\LaTeXe\ guide for the Journal of Functional Programming

### 2019-04-28 papers/jfp/4-schedulers.tex
Polish S4-S5

\noindent
As we see, the \cmd{main.exe} is started first as it is the required target. It
soon discovers a dependency on \cmd{util.o} and suspends to build \cmd{util.o},
resumes a build before building \cmd{main.o} then continues its work to finish
\cmd{main.exe}.

This scheduler (when combined with a suitable rebuilder) provides a minimal
build system which supports dynamic dependencies. In our model, a suspending
scheduler easy to write -- it makes a function call to compute each dependency. However,
a more practical implementation is likely to build multiple dependencies in parallel,
which then requires a more explicit suspension and resumption. To implement
suspension there are two standard approaches:

\begin{itemize}
\item Blocking threads or processes. This approach is relatively easy, but can
require significant resources, especially if a large number of tasks are
suspended. In languages with cheap green threads (e.g. Haskell) the approach is
more feasible, and it was the original approach taken by \Shake.
\item Continuation-passing style \cite{claessen_continuations} can allow the
remainder of a task to be captured, paused, and resumed later. Continuation
passing is efficient, but requires the build script to be architected to allow
capturing continuations. \Shake currently uses continuations.
\end{itemize}

>>>>

\noindent
As we see, \cmd{main.exe} is started first as it is the required target. It soon
discovers a dependency on \cmd{util.o} and suspends to build \cmd{util.o},
resumes a build before building \cmd{main.o} then continues its work to finish
\cmd{main.exe}.

This scheduler (when combined with a suitable rebuilder) provides a minimal
build system which supports dynamic dependencies. In our model, a suspending
scheduler is easy to write -- it makes a function call to compute each
dependency. However, a more practical implementation is likely to build multiple
dependencies in parallel, which then requires a more explicit task suspension
and resumption. To implement suspension there are two standard approaches:

\begin{itemize}
\item Blocking threads or processes. This approach is relatively easy, but can
require significant resources, especially if a large number of tasks are
suspended. In languages with cheap green threads (e.g. Haskell) the approach is
more feasible, and it was the original approach taken by \Shake.
\item Continuation-passing style~\cite{claessen_continuations} can allow the
remainder of a task to be captured, paused, and resumed later. Continuation
passing is efficient, but requires the build script to be architected to allow
capturing continuations. \Shake currently uses continuations.
\end{itemize}

### 2019-04-28 papers/jfp/5-rebuilders.tex
Polish S4-S5

A build system can be split into a scheduler (as defined
in~\S\ref{sec-scheduler}) and a \emph{rebuilder}. Suppose the scheduler decides
that a key should be brought up to date. The next question is: does any work
need to be done, or is the key already up to date? Or, in a cloud-build system,
do we have a cached copy of the value we need?

@@

More interestingly, to \emph{verify} whether a key needs rebuilding we use
\hs{verifyVT}, supplying the key, the hash of its current value, a function for
obtaining the post-build value of any key (using a scheduling strategy as
per~\S\ref{sec-scheduler}), and the existing trace store \hs{VT}. The result
will be a \hs{Bool} where \hs{True} indicates that the current value is already
up to date, and \hs{False} indicates that it should be rebuilt.

\q{Andrey}{The "function for obtaining the post-build value of any key" is a bit
mysterious. It would be nice to add a clarification. Or could we simplify the
interface and only provide a function with type
\hs{k -> Hash v -> [(k, Hash v)] -> VT k v -> Bool}? Presumably, \hs{verifyVT}
with the current type signature could be built on top of it?}

A verifying trace, and other types of traces discussed in this section, support
dynamic dependencies and minimality; furthermore, all traces except for deep
traces~(\S\ref{sec-deep-constructive-traces}) support early cutoff.

@@

The next optimisation is that there is only very minor benefit from storing more
than one \hs{Trace} per key. Therefore, verifying traces can be stored as
\hs{Map}~\hs{k}~\hs{(Trace}~\hs{k}~\hs{v}~\hs{(Hash}~\hs{v))}, where the
initial \hs{k} is the \hs{key} field of \hs{Trace}, thus making \hs{verifyVT}
fast. Note that storing only one \hs{Trace} per key means that if the
dependencies change but the result does not, and then the dependencies change
back to what they were before, there will be no valid \hs{Trace} available and
the build system will have to rebuild, whereas a complete list of all historical
traces would not have had to. However, it also means that the number of
\hs{Trace} structures is bounded by the number of distinct keys, regardless of
how many builds are executed -- a useful property.

@@

The \Shake build system, and the associated paper \cite{mitchell2012shake} in \S2.3.3 describes a slightly different trace structure, one that is similar to verifying traces, but stores slightly less data, and has slightly different early cutoff semantics. Rather than storing the \hs{Hash v} for each dependency, it instead stores a built time and changed time for each key, and for the dependencies just stores a list of \hs{k}. The resulting \hs{Trace} type ressembles:

\begin{minted}[fontsize=\small,xleftmargin=10pt]{haskell}
data Trace k v = Trace
    {key :: k
    ,result :: Hash v
    ,depends :: [k]
    ,built :: Time
    ,changed :: Time
    }
\end{minted}

The \hs{built} field is when the \hs{key} last rebuilt. The \hs{changed} field is when the \hs{result} last changed -- if the last build changed the value, it will be equal to \hs{built}, otherwise it will be older. The \hs{recordVT} function needs to consult the previous \hs{Trace} to know whether to use \hs{built} or the previous \hs{changed} value as the new \hs{changed} value. The \hs{verifyVT} needs to verify that the \hs{built} time of \hs{key} is newer than the \hs{changed} values of all its \hs{depends}. This scheme preserves minimality and early cutoff. A variant with only one \hs{Time} field would loose early cutoff, and indeed corresponds quite closely to \Make. Furthermore, the \hs{Time} stamp only needs record which execution of the build is running, so every key built in the same run can share the same \hs{Time} value -- it just needs to be monotically increasing between runs.

This optimisation is useful, at least in the case of \Shake, to save space. A hash takes up 32 bytes in typical programs, while a key (in \Shake only) is an \hs{Int} taking only 4 bytes. Furthermore, \Shake permits values to be arbitrarily large, and permits a custom equality (two values can be bit-for-bit unequal but considered equal by \Shake), meaning that a \hs{Hash} isn't a valid encoding. For build systems that used \hs{Applicative} dependencies, the \hs{depends} could be omitted entirely, making the size of a \hs{Trace} $O(1)$ instead of $O(n)$ where $n$ is the number of dependencies.

While verifying step traces are mostly an optimisation, there are some observable differences from verifying traces. Considering Figure \ref{fig-make}, imagine we built \hs{main.exe} (everything gets a \hs{built} and \hs{changed} of timestamp 1). Next we change \hs{util.c} and rebuild \hs{util.o} (both get a \hs{built} and \hs{changed} of timestamp 2). Next we put \hs{util.c} back to what it was originally, and build \hs{main.exe}. With verifying traces, the dependencies of \hs{main.exe} would be equal, and \hs{main.exe} would not rebuild. With verifying step traces, the \hs{changed} field of \hs{util.o} would increase once more, and \hs{main.exe} would rebuild. Other than when building subsets of the targets, we are unaware of any other situation where verifying step traces are less powerful.

>>>>

A build system can be split into a scheduler (as defined
in~\S\ref{sec-scheduler}) and a \emph{rebuilder}. Suppose the scheduler decides
that a key should be brought up to date. The next question is: does any work
need to be done, or is the key already up to date? Or, in a cloud build system,
do we have a cached copy of the value we need?

@@

More interestingly, to \emph{verify} whether a key needs rebuilding we use
\hs{verifyVT}, supplying the key, the hash of its current value, a function for
obtaining the hash of the post-build value of any key (using a scheduling
strategy as per~\S\ref{sec-scheduler}), and the existing trace store \hs{VT}.
The result will be a \hs{Bool} where \hs{True} indicates that the current value
is already up to date, and \hs{False} indicates that it should be rebuilt.

The most complex argument of \hs{verifyVT} is a function
\hs{fetchHash}~\hs{::}~\hs{k}~\hs{->}~\hs{m}~\hs{(Hash}~\hs{v)} to obtain the
hash of the post-build value of any key. With an applicative task,
\hs{fetchHash} will be called on the statically known task dependencies.
However, with a monadic task, the dependencies are not known from the task
alone, they are only recorded from previous executions and stored in \hs{VT}. If
the build system has two traces for a given key~\hs{k}, they will both request
the same dependency first, since \hs{Task}~\hs{Monad} is deterministic. However,
depending on that first result, they may then request different successive
dependencies using \hs{fetchHash}. A curious result is that for suspending
schedulers~(\S\ref{sec-suspending}) in many cases the actual build steps are
performed as a consequence of checking if a key needs rebuilding!

A verifying trace, and other types of traces discussed in this section, support
dynamic dependencies and minimality; furthermore, all traces except for deep
traces~(\S\ref{sec-deep-constructive-traces}) support the early cutoff
optimisation~(\S\ref{sec-background-shake}).

@@

The next optimisation is that there is only very minor benefit from storing more
than one \hs{Trace} per key. Therefore, verifying traces can be stored as
\hs{Map}~\hs{k}~\hs{(Trace}~\hs{k}~\hs{v}~\hs{(Hash}~\hs{v))}, where the
initial \hs{k} is the \hs{key} field of \hs{Trace}, thus making \hs{verifyVT}
much faster. Note that storing only one \hs{Trace} per key means that if the
dependencies of a key change but the resulting value does not, and then the
dependencies change back to what they were before, there will be no valid
\hs{Trace} available and the key will therefore have to be rebuilt, whereas a
complete list of all historical traces would allow the rebuilding to be skipped.
On the other hand, bounding the number of \hs{Trace} structures by the number of
distinct keys, regardless of how many builds are executed, is a useful property.

@@

The \Shake build system and the associated paper -- see~\S2.3.3
in~Mitchell~\shortcite{mitchell2012shake} -- use a slightly different trace
structure, called \emph{step traces}, one that is similar to verifying traces,
but stores less data, and has different early cutoff semantics. Rather than
storing the \hs{Hash}~\hs{v} for each dependency, it instead stores \hs{built}
time and \hs{changed} time for the \hs{key}, and a list of dependency keys
(without the hashes). The resulting \hs{StepTrace} type resembles:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
data StepTrace k v = StepTrace { key     :: k
                               , result  :: Hash v
                               , built   :: Time
                               , changed :: Time
                               , depends :: [k] }
\end{minted}
\vspace{1mm}

\noindent
The \hs{built} field is when the \hs{key} last rebuilt. The \hs{changed} field
is when the \hs{result} last changed -- if the last build changed the value, it
will be equal to \hs{built}, otherwise it will be older. The function
\hs{recordVT} consults the previous step traces to know whether to keep the
previous \hs{changed} value or change it to \hs{built}. The function
\hs{verifyVT} is a bit more subtle; given a key \hs{k} and the hash of its
current value \hs{h}, it performs the following steps:

\begin{itemize}
\item Find the latest (with respect to the field \hs{built}) step trace matching
  \hs{k}. If it does not exist, return \hs{False}: \hs{k} was never built
  before and cannot be verified.
\item If \hs{h} does not equal the \hs{result} field of the trace, return
  \hs{False}: the current \hs{k}'s value was written externally and needs
  rebuilding.
\item For each key \hs{d} in \hs{depends}:\\
  $\circ$ Make sure \hs{d} is up-to-date, suspending the current task if needed;\\
  $\circ$ If \hs{d}'s latest \hs{changed} time is greater than \hs{k}'s
    \hs{built} time, return \hs{False}.
\item Return \hs{True}: the current \hs{k}'s value is up-to-date.
\end{itemize}

\noindent
This approach preserves minimality and early cutoff. A variant with only one
\hs{Time} field would lose early cutoff, and indeed corresponds quite closely to
\Make. Furthermore, the \hs{Time} stamp only needs to record which execution of
the build is running, so every key built in the same run can share the same
\hs{Time} value -- it just needs to be monotonically increasing between runs.

This optimisation is useful, at least in the case of \Shake, to save space. A
typical cryptographic hash takes up 32 bytes, while a key (in \Shake only) is an
\hs{Int} taking only~4~bytes. Furthermore, \Shake permits values to be
arbitrarily large, and permits a custom equality (two values can be bit-for-bit
unequal but considered equal by \Shake), meaning that a \hs{Hash} is not a valid
encoding. For build systems that use \hs{Applicative} dependencies, the
\hs{depends} could be omitted entirely, making the size of a \hs{StepTrace}
$O(1)$ instead of $O(n)$, where $n$ is the number of dependencies.

\begin{figure}%[h]
\begin{subfigure}[b]{0.32\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/step-example-step1.pdf}}
\caption{Initial full build}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.32\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/step-example-step2.pdf}}
\caption{Change \cmd{util.c}, build \cmd{util.o}}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.33\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/step-example-step3.pdf}}
\caption{Restore \cmd{util.c}, build \cmd{main.exe}}
\end{subfigure}
% \vspace{-2mm}
\caption{An example of verifying step traces. The small rectangles show the
\hs{changed} (left) and \hs{built} (right) timestamps of each non-input key
in the trace store.
\label{fig-step}}
% \vspace{-2mm}
\end{figure}

While verifying step traces are mostly an optimisation, there are some
observable differences from verifying traces, as demonstrated by an example in
Fig.~\ref{fig-step}. We first make the full build: all keys get a \hs{built} and
\hs{changed} of timestamp 1. Next we change \cmd{util.c} and build \cmd{util.o};
the latter is changed as a result and hence both \hs{built} and \hs{changed} are
increased to 2. Finally, we change \cmd{util.c} back to what it was originally,
and build \cmd{main.exe}. With verifying traces, the hashes of the dependencies
of \cmd{main.exe} would be equal to the initial build, and \cmd{main.exe} would
not need rebuilding. With verifying step traces, the \hs{changed} field of
\cmd{util.o} would increase once more, and \cmd{main.exe} would therefore be
rebuilt. As shown in Fig.~\ref{fig-step}(c), the \hs{changed} field of
\cmd{main.exe} remains 1, since the actual value is unchanged. Other than when
building subsets of the targets, we are unaware of any other situation where
verifying step traces are less powerful.
